Investigating magnetic dynamo evolution with TESS field dwarfs

### 2017-10-05 ms.tex
more small tweaks

% Still to do: estimate the number of 16th mag stars and brighter in the whole
% sky.
% Estimate computation power available.
% Come up with a better target and priorities list.

@@

Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age, and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.
Angular momentum is carried away though magnetically driven stellar winds,
which slows the star's rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star's surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
Based on results from \Kepler, using precise light curves available from the
\TESS\ mission, we expect to measure the rotation periods of around 20\% of FG
stars and 80\% of KM stars in the \tess\ Full Frame Images (FFIs)\footnote{The
difference between these percentages comes from a dramatic increase in
magnetic activity, and therefore more easily measured rotation periods,
towards lower masses.}.
For many of these stars it will be possible to infer an age from their
rotation periods via gyrochronology.
Some outstanding mysteries regarding the nature of magnetic braking were
revealed, but thus-far unanswered, by \kepler.
Firstly, a mysterious gap in the rotation period distribution of \Kepler\
dwarfs requires either a sharp transition in magnetic dynamo geometry or a gap
in the local star formation history as an explanation \citep{mcquillan2014,
davenport2017}.
Secondly, a transitioning magnetic dynamo appears to be responsible for
inefficient magnetic braking at old ages in Solar-mass stars.
{\bf We propose to answer these questions by providing a catalog of
rotation periods for \TESS\ Full Frame Image (FFI) stars.}
% We also intend to use gyrochronology in the cases where it {\it is} applicable
% to learn about the evolution of exoplanets across the galaxy.

@@

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport found this period bimodality extends throughout all
masses in the Kepler rotation sample for nearby stars \citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The TESS FFI targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a discontinuous age distribution,
the feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the TESS FFI targets to the upcoming second data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
For the all-sky analysis we will focus on F and G stars since the 27.5 day
baseline of \TESS\ fields will limit measureable rotation periods to less than
around 15 days.
% With the April 2018 data release from Gaia (DR2), we estimate that we will be
% able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$3
% kpc.
% To characterize the rotation period bimodality as a function of galactic
% position, we propose to analyse the light curves of 10,000 FFI and two-minute
% targets that overlap with the Tycho-Gaia Astrometric Solution (TGAS) catalog,
% selecting FGKM dwarfs with a range of galactic positions.
% The FFI targets will be essential in order to reach $\sim$3kpc distances.
% Of these, the stars lying in the CVZs will be our highest priority.
The longer time coverage in the CVZz will allow us to study the rotation
period bimodality in K and M dwarfs, since the bimodality appears at longer
rotation periods for these lower mass stars.
{\bf Unlike \kepler\ and \ktwo\ light curves, \TESS\ provides the all-sky
coverage necessary to map the period bimodality across the galaxy.}

@@

Although the classical spin-down law of \citet{skumanich1972} (Period
$\propto$ Age$^1/2$) holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods, with the maximum
measurable period strongly depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the Continuous Viewing Zones (CVZs).
Many of these will have spectra from the \TESS --\HERMES\ survey, from which it
will be possible to calculate isochronal ages, enabling further calibration of
the gyrochronology relations.
We will also use the rotation periods of TESS field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many stars observable by \tess\ will have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action angles of the stars, both of which are age indicators.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.

In addition, we will use the rotation periods of comoving stars identified in
the first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
precision of gyrochronology.
We have identified 38 stars in 19 pairs with \Gaia\ proper motions and radial
velocities from \RAVE\ that are observable by \TESS, using {\tt tvguide}.
Of these, 34 stars in 17 pairs are in the CTL list.
Since these stars would make interesting targets for both asteroseismology and
exoplanet search, we include a list of them with this proposal.
The stars in these pairs are likely recently disrupted binaries so we
expect them to have the same age and measuring their rotation periods will
therefore provide a test of gyrochronology.

@@

\paragraph{Budget Justification}
PI Angus intends to use the budget to employ an undergraduate student for 3-4
months.
In this time the student will assist in extracting light curves from the
\tess\ FFIs, measuring rotation periods and building a rotation period
catalog.
The student will also be involved in the scientific project of their choosing:
either the rotation period bimodality, gyrochronology of low-mass stars, or
gyrochronology of comoving pairs.
The remaining budget will be used for traveling to meet with other Co-Is in
order to perform this analysis.
\paragraph{References}

>>>>

% Still to do: estimate the number of 16th mag stars and brighter in the TESS
% southern sky.
% Figure out how to incorporate the target list.

@@

Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.
Angular momentum is carried away though magnetically driven stellar winds,
which slows the star's rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star's surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
Based on results from \Kepler, using precise light curves available from the
\TESS\ mission, we expect to measure the rotation periods of around 20\% of FG
stars and 80\% of KM stars in the \tess\ Full Frame Images (FFIs)\footnote{The
difference between these percentages comes from a dramatic increase in
magnetic activity, and therefore more easily measured rotation periods,
towards lower masses.}.
For many of these stars it will be possible to infer an age from their
rotation periods via gyrochronology.
Some outstanding mysteries regarding the nature of magnetic braking were
revealed, but thus-far unanswered, by \kepler.
Firstly, a mysterious gap in the rotation period distribution of \Kepler\
dwarfs requires either a sharp transition in magnetic dynamo geometry or a gap
in the local star formation history as an explanation \citep{mcquillan2014,
davenport2017}.
Secondly, a transitioning magnetic dynamo appears to be responsible for
inefficient magnetic braking at old ages in Solar-mass stars.
To answer these questions {\bf we propose to provide a catalog of rotation
periods for \TESS\ Full Frame Image (FFI) stars.}

@@

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport found this period bimodality extends throughout all
masses in the Kepler rotation sample for nearby stars \citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The TESS FFI targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a discontinuous age distribution,
the feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the TESS FFI targets to the upcoming second data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
For the all-southern-hemisphere  analysis we will focus on F and G stars since
the 27.4 day baseline of \TESS\ fields will limit measureable rotation periods
to less than around 15 days.
The longer time coverage in the CVZz will allow us to study the rotation
period bimodality in K and M dwarfs, since the bimodality appears at longer
rotation periods for these lower mass stars.
Unlike \kepler\ and \ktwo\ light curves, {\bf \TESS\ provides the all-sky
coverage necessary to map the period bimodality across the galaxy.}

@@

Although the classical spin-down law of \citet{skumanich1972} (Period
$\propto$ Age$^1/2$) holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods, with the maximum
measurable period strongly depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the Continuous Viewing Zones (CVZs).
Many of these will have spectra from the \TESS --\HERMES\ survey, from which
it will be possible to calculate isochronal ages, enabling further calibration
of the gyrochronology relations.
{\bf We propose to calibrate gyrochronology at old ages for FGKM stars using
the isochronal ages of \TESS--\hermes\ stars.}
We will also use the rotation periods of TESS field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many stars observable by \tess\ will have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action angles of the stars, both of which are age indicators.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
{\bf We propose to calibrate the gyrochronology relations at old ages for FGKM
stars using vertical actions of \Gaia\ DR2 targets in the CVZ.}

We have identified 34 stars in 17 pairs with \Gaia\ proper motions and radial
velocities from \RAVE\ that are observable by \TESS
% \footnote{Based on {\tt tvguide}.}
and are already in the Candidate Target List.
Since these stars would make interesting targets for both asteroseismology and
exoplanet search, we include a list of them with this proposal.
The stars in these pairs are likely recently disrupted binaries so we
expect them to have similar ages and measuring their rotation periods will
therefore provide a test of gyrochronology.
{\bf We will use the rotation periods of comoving stars identified in the
first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
precision of gyrochronology.}

@@

\paragraph{Budget Justification}
PI Angus intends to use the budget to employ an undergraduate student for 3-4
months.
In this time the student will assist in extracting light curves from the
\tess\ FFIs, measuring rotation periods and building a rotation period
catalog.
The student will also be involved in the scientific project of their choosing:
either the rotation period bimodality, gyrochronology of FGKM stars, or
gyrochronology of comoving pairs.
The remaining budget will be used for traveling to meet with other Co-Is in
order to perform this analysis.
\paragraph{References}
